### 2025-07-10 manuscript_ten-simple-rules-writing-retreat.tex
removed double brackets around citations

At every stage in a researcher's career, scholarly output advances scientific knowledge and supports career development. Early career researchers, in particular, significantly boost their career prospects by increasing their scholarly outputs [~\cite{horta2016, wilkins2021}]. Writing serves as an integral skill for academic work [~\cite{boice1984, hazelett2025}], especially when competing for grants and jobs. Academics juggle administrative tasks alongside teaching, collection and analysis of data, and production of publications and presentations. Consequently, many report a lack of time to think critically as a major challenge in academia [~\cite{menzies2007}], which too often leads researchers to deprioritise writing tasks [~\cite{boice1984, menzies2007}]. Therefore, we must actively disengage from other tasks to fully engage with the act of writing [~\cite{murray2009, murray2013}].

Writing retreats provide structured periods where academics dedicate time to focused writing [~\cite{mcgrail2006, murray2009}]. These retreats offer practical opportunities to disconnect from daily work routines [~\cite{murray2013, tremblay2021}], which help researchers gain writing momentum and increase scholarly output [~\cite{mcgrail2006, kornhaber2016}]. In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers [~\cite{kornhaber2016, eardley2021, tremblay2021}]. Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing [~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}].

As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023. In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, and each student conducts research individually. In addition, given our diverse disciplines, field sites, and lab locations, we had felt a disconnection among us and a lack of community. Therefore, we initially planned the retreat to connect with our PhD cohort [~\cite{omeara2017, bernery2022}], but we also aimed to make it a productive work week [~\cite{mcgrail2006, kornhaber2016}]. After returning from the retreat, we realised how much writing we had accomplished and the importance of the moments we had shared as postgrads going through similar experiences in various stages of our PhDs. The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.

Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats. Although the authors were all PhD students at the time of the retreats, these rules apply broadly to any research-oriented or academically-minded group. We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se. Many resources on academic writing exist that we encourage you to explore [e.g.~\cite{turbek2016, hotaling2020, carter2020, cargill2021}], including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]---or you may simply choose to write a boring contribution instead [~\cite{sand2007}].

@@

A change of scenery inspires productivity [~\cite{murray2013}], whether you organise the writing retreat locally or as a multi-day trip. Local facilities, such as meeting rooms and other bookable spaces on campus, offer accessible options. For longer retreats, students---regardless of departmental affiliation---should inquire about their university's field stations. For example, many biology and ecology programmes maintain field stations (for a global list, see [~\cite{tydecks2016}]). By using university-owned facilities, students keep retreat costs manageable. Field stations often remain underutilized during certain periods, so retreats during these times can be used to promote year-round upkeep of facilities. In addition, field stations often sit in remote, natural settings with limited distractions, creating ideal environments for focus and uninterrupted work [~\cite{murray2013, tremblay2021}]. If your university lacks a suitable field station, other institutions may allow shared use depending on availability. We encourage students to contact station managers and build those connections.

@@

Create a schedule to structure your writing retreat, as this will be key to its success [~\cite{murray2009, tremblay2021}]. Use available resources to guide your planning [e.g. ~\cite{tremblay2021}], including this guide (see also Appendices). At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats. We encourage you to take advantage of such services when available.

@@

Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions [~\cite{tremblay2021}; see also \nameref{S1_Appendix}]. We designated a different facilitator each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes). Each morning began with a 15-minute writing session using prompts that ranged from reflective writing to humorous research titles (see Rule 9). These sessions helped rev up our writing engines for the day and offered a fun, low-pressure opportunity to share writing. Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner' [see Rule 5 in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).

@@

We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group. Goals may be quantitative (e.g. word count, page number) or qualitative (e.g. specific sections or revisions). For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts. Identifying a specific writing target ahead of time reduces decision-making friction at the retreat and maximises productive preparation. Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.
Hold the pre-retreat meeting within two weeks of departure. This timeframe supports goal-oriented preparation while preserving participant momentum. If possible, host the meeting in person in a casual setting. If attendees join from different campuses, a video call works as well. Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g. meal coordination, daily facilitator), or ideas for social activities.

@@

However, writing is not easy, and writers face many hurdles, such as a lack of motivation, uncertainty about what to do next, writer's block, imposter syndrome, and more [~\cite{boice1984, murray2013, wilson2019, grogan2021}]. These obstacles can easily disrupt even the best-laid plans and most robust retreat structures. Therefore, have a backup plan. If it becomes challenging to write, consider taking a break or switching tasks; it is proven that even short breaks increase productivity [~\cite{carter2020, lyubykh2022}]. Everyone has different strategies that work for them. For example, some people prefer to cite as they write, while others write everything out quickly and add citations during editing. Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block [~\cite{hemingway1935}]. Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you [~\cite{peterson2018, grogan2021}].

@@

Undertaking any large-writing project (e.g. a PhD thesis) is a daunting and varying experience for everyone. It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation [~\cite{wilson2019}]. A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers. While we found this occurred naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees. The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.'

We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat. Access to a projector and screen in our second and third retreats was beneficial. However, previously, we have simply shared from our laptop screens or consulted notes and found it equally engaging due to the small group attending the retreats. Discussions were informal, and we aimed not to exceed 20 minutes per session. It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation. These are simply opportunities to share, learn, and exchange insights. Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students [~\cite{omeara2017, bernery2022}].

@@

Schedule time to exchange written material and review that of your peers. An essential aspect of developing your writing is giving and receiving revisions [~\cite{guilford2001, reynolds2011}]. Constructive feedback from peers may offer valuable insights on gaps in clarity, logic, and structure. Even if you do not understand the scientific content of the work, focusing on structure, prose, and writing flow can be extremely helpful. This type of review can unblock a colleague who feels stuck. Likewise, feedback from a peer outside one's field may offer perspectives the author had not previously considered.

Providing feedback is just as valuable as receiving it. This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed. When giving feedback, use neutral language [~\cite{parsons2021}] and focus on writing components (structure, clarity, flow). If you are an expert in the content, detailed feedback can also be extremely valuable. Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers [~\cite{smith2023, saleh2024}], and work towards breaking this pattern. Furthermore, peer reviewing should be an integral part of academic life [~\cite{borja2024}]. Recognise peer review as a skill that can be practised, and a writing retreat could be the perfect place to begin if you are new to it.

@@

Have fun and build connections with your peers. While on a writing retreat, including recreational activities can enhance the overall experience. A writing retreat allows you to disconnect from the routine and responsibilities [~\cite{murray2013}], providing a unique opportunity to bond with your peers. More rural places often have scenic landscapes, hospitable communities and limited distractions. While this may restrict choices for activities, it can lead to higher quality experiences where you can get to know local businesses and each other on a more personal level. Before embarking on a writing retreat, research what is unique about the location and take advantage of what may be on offer. We organised creative writing prompts for each morning (see Rule 3) and gave casual presentations (Rule 7) about various topics, which were fun and engaging bonding activities. Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity [~\cite{eardley2021, lyubykh2022}], enhancing the writing experience.

@@

Collectively, we experienced myriad benefits of writing retreats reported in the literature. These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara2017}] and decreasing our sense of isolation [~\cite{eardley2021}]. Most importantly, we achieved our writing goals at the retreats. Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time [~\cite{peterson2018, grogan2021}]. By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts. Happy writing!

>>>>

At every stage in a researcher's career, scholarly output advances scientific knowledge and supports career development. Early career researchers, in particular, significantly boost their career prospects by increasing their scholarly outputs ~\cite{horta2016, wilkins2021}. Writing serves as an integral skill for academic work ~\cite{boice1984, hazelett2025}, especially when competing for grants and jobs. Academics juggle administrative tasks alongside teaching, collection and analysis of data, and production of publications and presentations. Consequently, many report a lack of time to think critically as a major challenge in academia ~\cite{menzies2007}, which too often leads researchers to deprioritise writing tasks ~\cite{boice1984, menzies2007}. Therefore, we must actively disengage from other tasks to fully engage with the act of writing ~\cite{murray2009, murray2013}.

Writing retreats provide structured periods where academics dedicate time to focused writing ~\cite{mcgrail2006, murray2009}. These retreats offer practical opportunities to disconnect from daily work routines ~\cite{murray2013, tremblay2021}, which help researchers gain writing momentum and increase scholarly output ~\cite{mcgrail2006, kornhaber2016}. In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers ~\cite{kornhaber2016, eardley2021, tremblay2021}. Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing ~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}.

As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023. In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, and each student conducts research individually. In addition, given our diverse disciplines, field sites, and lab locations, we had felt a disconnection among us and a lack of community. Therefore, we initially planned the retreat to connect with our PhD cohort ~\cite{omeara2017, bernery2022}, but we also aimed to make it a productive work week ~\cite{mcgrail2006, kornhaber2016}. After returning from the retreat, we realised how much writing we had accomplished and the importance of the moments we had shared as postgrads going through similar experiences in various stages of our PhDs. The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.

Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats. Although the authors were all PhD students at the time of the retreats, these rules apply broadly to any research-oriented or academically-minded group. We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se. Many resources on academic writing exist that we encourage you to explore (e.g. ~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}---or you may simply choose to write a boring contribution instead ~\cite{sand2007}.

@@

A change of scenery inspires productivity ~\cite{murray2013}, whether you organise the writing retreat locally or as a multi-day trip. Local facilities, such as meeting rooms and other bookable spaces on campus, offer accessible options. For longer retreats, students---regardless of departmental affiliation---should inquire about their university's field stations. For example, many biology and ecology programmes maintain field stations (for a global list, see ~\cite{tydecks2016}). By using university-owned facilities, students keep retreat costs manageable. Field stations often remain underutilized during certain periods, so retreats during these times can be used to promote year-round upkeep of facilities. In addition, field stations often sit in remote, natural settings with limited distractions, creating ideal environments for focus and uninterrupted work ~\cite{murray2013, tremblay2021}. If your university lacks a suitable field station, other institutions may allow shared use depending on availability. We encourage students to contact station managers and build those connections.

@@

Create a schedule to structure your writing retreat, as this will be key to its success ~\cite{murray2009, tremblay2021}. Use available resources to guide your planning (e.g. ~\cite{tremblay2021}), including this guide (see also Appendices). At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats. We encourage you to take advantage of such services when available.

@@

Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions ~\cite{tremblay2021} (see also \nameref{S1_Appendix}). We designated a different facilitator each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes). Each morning began with a 15-minute writing session using prompts that ranged from reflective writing to humorous research titles (see Rule 9). These sessions helped rev up our writing engines for the day and offered a fun, low-pressure opportunity to share writing. Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner'; see Rule 5 in ~\cite{peterson2018}), a shared dinner, and presentations (see Rule 7).

@@

We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group. Goals may be quantitative (e.g. word count, page number) or qualitative (e.g. specific sections or revisions). For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts. Identifying a specific writing target ahead of time reduces decision-making friction at the retreat and maximises productive preparation. Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.

Hold the pre-retreat meeting within two weeks of departure. This timeframe supports goal-oriented preparation while preserving participant momentum. If possible, host the meeting in person in a casual setting. If attendees join from different campuses, a video call works as well. Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g. meal coordination, daily facilitator), or ideas for social activities.

@@

However, writing is not easy, and writers face many hurdles, such as a lack of motivation, uncertainty about what to do next, writer's block, imposter syndrome, and more ~\cite{boice1984, murray2013, wilson2019, grogan2021}. These obstacles can easily disrupt even the best-laid plans and most robust retreat structures. Therefore, have a backup plan. If it becomes challenging to write, consider taking a break or switching tasks; it is proven that even short breaks increase productivity ~\cite{carter2020, lyubykh2022}. Everyone has different strategies that work for them. For example, some people prefer to cite as they write, while others write everything out quickly and add citations during editing. Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block ~\cite{hemingway1935}. Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you ~\cite{peterson2018, grogan2021}.

@@

Undertaking any large-writing project (e.g. a PhD thesis) is a daunting and varying experience for everyone. It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}. A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers. While we found this occurred naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees. The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.'

We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat. Access to a projector and screen in our second and third retreats was beneficial. However, previously, we have simply shared from our laptop screens or consulted notes and found it equally engaging due to the small group attending the retreats. Discussions were informal, and we aimed not to exceed 20 minutes per session. It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation. These are simply opportunities to share, learn, and exchange insights. Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students ~\cite{omeara2017, bernery2022}.

@@

Schedule time to exchange written material and review that of your peers. An essential aspect of developing your writing is giving and receiving revisions ~\cite{guilford2001, reynolds2011}. Constructive feedback from peers may offer valuable insights on gaps in clarity, logic, and structure. Even if you do not understand the scientific content of the work, focusing on structure, prose, and writing flow can be extremely helpful. This type of review can unblock a colleague who feels stuck. Likewise, feedback from a peer outside one's field may offer perspectives the author had not previously considered.

Providing feedback is just as valuable as receiving it. This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed. When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components (structure, clarity, flow). If you are an expert in the content, detailed feedback can also be extremely valuable. Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers ~\cite{smith2023, saleh2024}, and work towards breaking this pattern. Furthermore, peer reviewing should be an integral part of academic life ~\cite{borja2024}. Recognise peer review as a skill that can be practised, and a writing retreat could be the perfect place to begin if you are new to it.

@@

Have fun and build connections with your peers. While on a writing retreat, including recreational activities can enhance the overall experience. A writing retreat allows you to disconnect from the routine and responsibilities ~\cite{murray2013}, providing a unique opportunity to bond with your peers. More rural places often have scenic landscapes, hospitable communities and limited distractions. While this may restrict choices for activities, it can lead to higher quality experiences where you can get to know local businesses and each other on a more personal level. Before embarking on a writing retreat, research what is unique about the location and take advantage of what may be on offer. We organised creative writing prompts for each morning (see Rule 3) and gave casual presentations (Rule 7) about various topics, which were fun and engaging bonding activities. Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity ~\cite{eardley2021, lyubykh2022}, enhancing the writing experience.

@@

Collectively, we experienced myriad benefits of writing retreats reported in the literature. These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment ~\cite{papen2018}, increasing our sense of belonging as PhD students ~\cite{omeara2017} and decreasing our sense of isolation ~\cite{eardley2021}. Most importantly, we achieved our writing goals at the retreats. Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time ~\cite{peterson2018, grogan2021}. By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts. Happy writing!
